---
title: "EC4308 Project (Lending Club)"
author: "Brandon, LX, WT, YH, ZH"
date: "2024-10-11"
output: html_document
---

## Initial setup
```{r setup, include=FALSE}
library(readr)
library(tidyverse)

lc_loan <- read_csv("data/lc_loan.csv")
lc_2016_2017 <- read.csv("data/lc_2016_2017.csv")
```

### Analysing the head of the data
```{r cars}
head(lc_loan)
```

## Data Cleaning
### Observe the difference in columns
```{r}
# Get the column names from both dataframes
lc_loan_cols <- colnames(lc_loan)
lc_2016_2017_cols <- colnames(lc_2016_2017)

# Compare the columns to find the differences
diff_cols <- setdiff(lc_loan_cols, lc_2016_2017_cols)
diff_cols_2 <- setdiff(lc_2016_2017_cols, lc_loan_cols)

# Display the differences
cat("Columns in 'lc_loan' but not in 'lc_2016_2017':", diff_cols, "\n")
cat("Columns in 'lc_2016_2017' but not in 'lc_loan':", diff_cols_2, "\n")
```

### Remove columns that are not common between the two dataframes
```{r}
lc_loan_clean <- lc_loan %>% select(-all_of(diff_cols))
lc_2016_2017_clean <- lc_2016_2017 %>% select(-all_of(diff_cols_2))
combined_data <- bind_rows(lc_loan_clean, lc_2016_2017_clean)
glimpse(combined_data)
```

## WT + ZH
## Index loan status to 0 = Fully paid + Current, 1 = Late + Charged Off + Late  + Grace period + Default
### Convert loan_status to a factor (categorical variable)
```{r}
combined_data <- combined_data %>%
  mutate(loan_status = as.factor(loan_status))
table(combined_data$loan_status)
```


### Print out the possible levels of loan_status
```{r}
cat("Possible loan_status categories:\n")
print(levels(combined_data$loan_status))
```

### Now convert it to binary 
```{r}
combined_data <- combined_data %>%
  mutate(
    loan_status_bin = case_when(
      loan_status %in% c("Current", "Fully Paid", "Does not meet the credit policy. Status:Fully Paid") ~ 0,
      TRUE ~ 1  # Everything else will be 1
    )
  )

head(combined_data)
```

### Check how many 0s and 1s in the new binary column
```{r}
table(combined_data$loan_status_bin)
```
~ 8.5% are 1

### Stratify the data based on the loan_status and get ~ 50,000 accordingly
#### Rdm suggestion, compare the direct pulling of 50k data points vs stratified (Prof might be interested)
```{r}
# Set the seed for reproducibility
set.seed(4308)

# Randomly sample 50,000 rows from the dataset
random_sampled_data <- combined_data %>%
  sample_n(size = 50000)

# Check the distribution of loan_status_bin in the sampled data
loan_status_proportion <- table(random_sampled_data$loan_status_bin)

# Display the proportions
loan_status_proportion
```
~8.4% are 1 as well, nice.


### Label the data


# Brandon + LX
## Include Fed data Last payment date + 3 months

# YH + Grab anyone for help
## Index term to 0 or 1
## Drop grade Adds bias
## Index emp_length -> <1 year = 0
## Replace NA with 0

# Once all is done
## 70/30 Data split
## Save CSV file as combined_data_cleaned.csv

# Future plans
## EDA
## Scale the data for the relevant regressions
## If the output is a probability, then >0.5 = bad loan, else good loan

# Models to be use
## 1. Benchmark model = Decision Tree


